Investigating magnetic dynamo evolution with \TESS\ field dwarfs

### 2017-10-06 ms.tex
estimating number of rotation periods

% Still to do: estimate the number of 16th mag stars and brighter in the TESS
% southern sky. ~ 15 million, call it 10.
% Cite the sullivan paper.
% Figure out how to incorporate the target list.
% Get detection probabilities for stars on the list
% Calculate the distance at which you get a ~ 15th mag at ~1kpc
% Mention vetting.

@@

Of the measurable properties for a large ensemble of field stars, rotation
periods contain the most information about stellar age and provide the best
leverage for advancing our knowledge of galactic archeology as well as
exoplanet population demographics via gyrochronology.
Angular momentum is carried away though magnetically driven stellar winds,
which slows the star's rotation over cosmic time.
This rotation-based `clock' is known as gyrochronology.
Cool spots on the star's surface rotate in-to and out-of view, creating small
amplitude ($\pm$ 1\% or less) quasi-periodic changes in the stellar
brightness.
Based on results from \Kepler, using precise light curves available from the
\TESS\ mission, we expect to measure the rotation periods of around 20\% of FG
stars and 80\% of KM stars in the \tess\ Full Frame Images (FFIs)\footnote{The
difference between these percentages comes from a dramatic increase in
magnetic activity, and therefore more easily measured rotation periods,
towards lower masses.}.
For many of these stars it will be possible to infer an age from their
rotation periods via gyrochronology.
Some outstanding mysteries regarding the nature of magnetic braking were
revealed, but thus-far unanswered, by \kepler.
Firstly, a mysterious gap in the rotation period distribution of \Kepler\
dwarfs requires either a sharp transition in magnetic dynamo geometry or a gap
in the local star formation history as an explanation \citep{mcquillan2014,
davenport2017}.
Secondly, a transitioning magnetic dynamo appears to be responsible for
inefficient magnetic braking at old ages in Solar-mass stars
\citep{vansaders2016}.
To answer these questions {\bf we propose to provide a catalog of rotation
periods for \TESS\ Full Frame Image (FFI) stars.}

@@

Although the classical spin-down law of \citet{skumanich1972} (Period
$\propto$ Age$^{1/2}$) holds for all open clusters with measured periods, it does
not appear to describe old field stars \citep{angus2015, van-saders2016}.
\citet{van-saders2016} found that including a transition to a weakened magnetic
braking regime in the gyrochronology models, at a critical Rossby number,
provides an improved fit to the data.
Further calibration is needed however; a lack of rotation periods and reliable
ages for old and low-mass stars leaves the rotational evolution of stars older
than the Sun relatively unexplored.
\TESS\ will provide thousands of new rotation periods, with the maximum
measurable period strongly depending on the length of time it is observed for.
Of particular interest are the potentially long rotation periods measurable
for stars that fall in the CVZ.
Many of these will have spectra from the \TESS --\HERMES\ survey, from which
it will be possible to calculate isochronal ages, enabling further calibration
of the gyrochronology relations.
{\bf We propose to calibrate gyrochronology at old ages for FGKM stars using
the isochronal ages of \TESS--\hermes\ stars.}
We will also use the rotation periods of \TESS\ field dwarfs over the whole sky
to test the gyrochronology relations using galactic kinematics.
Many stars observable by \tess\ will have proper motions, parallaxes,
positions and radial velocities published in the second \Gaia\ data release.
These parameters provide the information necessary to calculate galactocentric
positions and action-angle coordinates of the stars, both of which are age
indicators.
Since one of the few ways to accurately age-date fully convective, late M
dwarfs is via kinematics, these new relations will help to infer ages for all
stars with M$<$ 0.35M$_\odot$, to which gyrochronology cannot be applied.
{\bf We propose to calibrate the gyrochronology relations at old ages for FGKM
stars using vertical actions of \Gaia\ DR2 targets in the CVZ.}

@@

Based on \citet{sullivan2015}, we expect around 10 million \TESS\ FFI
targets brighter than 15th magnitude in the southern ecliptic hemisphere.
We intend to prioritize certain groups of stars for period analysis,
calculating a simple (and computationally fast) Lomb-Scargle periodogram for
the low priority objects and running a full probabilistic, Gaussian process
period analysis on the high priority targets.
High priority targets will include: comoving pairs, stars in the CVZ
(especially those with spectra), stars in TGAS, stars at large distances, plus
any potential asteroseismic dwarfs and exoplanet hosts.
Based on the expected photometric precision of \tess\ \citep{sullivan2015}, we
expect to be able to measure rotation periods for FFI stars down to 15th
magnitude.
We note, however that not all stars will have measurable rotation periods,
even if they are bright enough --- not all stars show rotational variability
in their light curves either because they are pole-on or because they are
inactive.
We will vet incorrect rotation period measurements using a combination of a
Lomb-Scargle periodogram and an autocorrelation function by
training on a set of light curves with a range of variability amplitudes and
periods.
However, based on the high success rates of a Lomb-Scargle periodogram method
with high-pass filtered light curve and the Gaussian process method
demonstrated in \citet{angus2017}, we expect a low rate of incorrect rotation
periods.

@@

\paragraph{Expected Impact}
We will provide rotation period posterior probability density functions and
best fit estimates for around 30,000 high priority FFI targets rotation
periods point estimates for lower priority FFI stars using a combination of
Lomb-Scargle periodogram and autocorrelation techniques.
% Finally, we will release a catalog of rotation periods for all two-minute
% cadence targets.
The full catalog will contain several million rotation periods.
We will release code for inferring stellar ages from rotation periods, based
on a new gyrochronology relation calibrated using \TESS -- \HERMES spectra,
\Gaia\ kinematics and comoving pairs.
We will also release an open source python package for measuring rotation
periods of \tess\ FFI targets.

\paragraph{Budget Justification}
PI Angus intends to use the budget to employ an undergraduate student for 3-4
months.
In this time the student will assist in extracting light curves from the
\tess\ FFIs, measuring rotation periods and building a rotation period
catalog.
The student will also be involved in the scientific project of their choosing:
either the rotation period bimodality, gyrochronology of FGKM stars, or
gyrochronology of comoving pairs.
The remaining budget will be used for traveling to meet with other Co-Is in
order to perform this analysis.
\paragraph{References}

>>>>

% Figure out how to incorporate the target list.
% Get detection probabilities for stars on the list
% Calculate the distance at which you get a ~ 15th mag at ~1kpc
% Shorten.

@@

Of the measurable properties for a large ensemble of field stars, rotation
periods contain the most information about stellar age and provide the best
leverage for advancing our knowledge of galactic archeology as well as
exoplanet population demographics via gyrochronology.
Angular momentum is carried away though magnetically driven stellar winds,
which slows the star's rotation over cosmic time.
This rotation-based `clock' is known as gyrochronology.
Cool spots on the star's surface rotate in-to and out-of view, creating small
amplitude ($\pm$ 1\% or less) quasi-periodic changes in the stellar
brightness.
Based on results from \Kepler, using precise light curves available from the
\TESS\ mission, we expect to measure the rotation periods of around 20\% of FG
stars and 80\% of KM stars in the \tess\ Full Frame Images (FFIs)\footnote{The
difference between these percentages comes from a dramatic increase in
magnetic activity, and therefore more easily measured rotation periods,
towards lower masses.}.
For many of these stars it will be possible to infer an age from their
rotation periods via gyrochronology.
Some outstanding mysteries regarding the nature of magnetic braking were
revealed, but thus-far unanswered, by \kepler.
Firstly, a mysterious gap in the rotation period distribution of \Kepler\
dwarfs requires either a sharp transition in magnetic dynamo geometry or a gap
in the local star formation history as an explanation \citep{mcquillan2014,
davenport2017}.
Secondly, a transitioning magnetic dynamo appears to be responsible for
inefficient magnetic braking at old ages in Solar-mass stars
\citep{van-saders2016}.
To answer these questions {\bf we propose to provide a catalog of rotation
periods for \TESS\ Full Frame Image (FFI) stars.}

@@

Although the classical spin-down law of \citet{skumanich1972} (Period
$\propto$ Age$^{1/2}$) holds for all open clusters with measured periods, it does
not appear to describe old field stars \citep{angus2015, van-saders2016}.
\citet{van-saders2016} found that including a transition to a weakened magnetic
braking regime in the gyrochronology models, at a critical Rossby number,
provides an improved fit to the data.
Further calibration is needed however; a lack of rotation periods and reliable
ages for old and low-mass stars leaves the rotational evolution of stars older
than the Sun relatively unexplored.
\TESS\ will provide thousands of new rotation periods, with the maximum
measurable period strongly depending on the length of time it is observed for.
Of particular interest are the potentially long rotation periods measurable
for stars that fall in the CVZ.
Many of these will have spectra from the \TESS --\HERMES\ survey, from which
it will be possible to calculate isochronal ages, enabling further calibration
of the gyrochronology relations.
{\bf We propose to calibrate gyrochronology at old ages for FGKM stars using
the isochronal ages of \TESS--\hermes\ stars.}
We will also use the rotation periods of \TESS\ field dwarfs over the whole sky
to test the gyrochronology relations using galactic kinematics.
Many \tess\ FFI targets will have proper motions, parallaxes, positions and
radial velocities published in the second \Gaia\ data release.
These parameters provide the information necessary to calculate galactocentric
positions and action-angle coordinates of the stars, both of which are age
indicators.
Since one of the few ways to accurately age-date fully convective, late M
dwarfs is via kinematics, these new relations will help to infer ages for all
stars with M$<$ 0.35M$_\odot$, to which gyrochronology cannot be applied.
{\bf We propose to calibrate the gyrochronology relations at old ages for FGKM
stars using vertical actions of \Gaia\ DR2 targets in the CVZ.}

@@

Based on \citet{sullivan2015}, we expect around 10 million \TESS\ FFI
targets brighter than 15th magnitude in the southern ecliptic hemisphere.
We intend to prioritize certain groups of stars for period analysis,
calculating a simple (and computationally fast) Lomb-Scargle periodogram for
the low priority objects and running a full probabilistic, Gaussian process
period analysis on the high priority targets.
High priority targets will include: comoving pairs, stars in the CVZ
(especially those with spectra), stars in TGAS, stars at large distances, plus
any potential asteroseismic dwarfs and exoplanet hosts.
The number of high priority targets with probabilistic rotation periods will
be limited by computation power, but we conservatively estimate that we can
produce full rotation period posterior PDFs for $\sim$ 30,000 stars.
Based on the expected photometric precision of \tess\ \citep{sullivan2015}, we
expect to be able to measure rotation periods for FFI stars down to 15th
magnitude.
We note, however that not all stars will have measurable rotation periods,
even if they are bright enough not all stars show rotational variability in
their light curves either because they are pole-on or inactive.
We will vet incorrect rotation period measurements using a combination of a
Lomb-Scargle periodogram and an autocorrelation function by
training on a set of light curves with a range of variability amplitudes and
periods.
However, based on the high success rates of a Lomb-Scargle periodogram method
with high-pass filtered light curve and the Gaussian process method,
demonstrated in \citet{angus2017}, we expect a low rate of incorrect rotation
periods.

@@

\paragraph{Expected Impact}
We will provide rotation period posterior probability density functions and
best fit estimates for $\sim$ 30,000 high priority FFI targets rotation
periods and point estimates for $\sim$ 10 million lower priority FFI stars
using a combination of Lomb-Scargle periodogram and autocorrelation
techniques.
We will release an open source python package for measuring rotation periods
of \tess\ FFI targets plus code for inferring stellar ages from rotation
periods, based on a new gyrochronology relation calibrated using \TESS --
\HERMES spectra, \Gaia\ kinematics and comoving pairs.

\paragraph{Budget Justification}
PI Angus will employ two undergraduate students for 3-4 months who will assist
in extracting light curves from the \tess\ FFIs, measuring rotation periods
and building a rotation period catalog.
These students will also be involved in the scientific project of their
choosing: either the rotation period bimodality, gyrochronology of FGKM stars,
or gyrochronology of comoving pairs.
The remaining budget will be used for a bringing all Co-Is together in order
to make and execute analysis plans and for presenting the results at
conferences.
\paragraph{References}
